Verification of Stencil Computations through Spatial Specifications

### 2016-07-07 paper.tex
intro edits

\keywords{program verification, specifications, static analysis,
  stencil computation}

@@

In practice, most stencil computations have a regular shape that can
be described simply and abstractly with a small set of coarse-grained
spatial descriptions. We introduce a simple specification language of
such descriptions, abstracting over the fine-grained detail of stencil
access patterns. For our first Laplace example, our
inference procedure provides the following specification to line 3:
%
\begin{minted}[linenos=false]{fortran}
!=  stencil centered(depth=1, dim=1) :: a
\end{minted}
%
This explains that \texttt{a} is read from with a
symmetrical access pattern (``centered'') to a depth of one in each
direction in its first (and only) dimension.
%The second line explains
%the temporal relationship between \texttt{b} and \texttt{a}: that
%previous time step for \texttt{b} is actually provided by \texttt{a},
%and vice versa. This is explained as a mutual dependence between
%\texttt{a} and \texttt{b}.
In the case of the Navier-Stokes example of
Figure~\ref{ref:navier-stokes-fragment}, its inferred specification is shown
in Figure~\ref{ref:navier-stokes-fragment}(b). The
specification explains that, over the whole fragment, \texttt{u} is
read from with a centered pattern to depth of 1 in both dimensions
(this is known as the \emph{five point stencil}) and \texttt{v}
is read from in a neighbourhood bounded by forward to depth of $1$ in
the first dimension and backward to a depth of $1$ in the second
dimension. Figure~\ref{ref:navier-stokes-fragment}(c) illustrates
the specifications pictorially.

In this paper, we make the following contributions:
%
\begin{itemize}
\item we introduce our specification language for
stencil computations that captures many common forms
of data access pattern (Section~\ref{sec:lang})
along with a sound denotational model (Section~\ref{sec:semantics});

\item we detail inference, checking, specification synthesis,
and program synthesis algorithms for our language
(Section~\ref{sec:analysis}), which relies on the semantic model;

\item we evaluate our approach, providing an
implementation as an extension of CamFort,
an open-source program analysis and refactoring tool for Fortran
corpus. We use this implementation to evaluate our approach against
a corpus of scientific applications, totalling roughly one million
lines of code, and we study some verification case studies;

@@

\item we consider two particular stencils in detail and show that we
  are able to detect the vast majority of (synthesised) possible
  indexing errors.
\end{itemize}
%

@@

duvdy = ((v(i,j)+v(i+1,j))*(u(i,j)+u(i,j+1))+ &
  gamma*abs(v(i,j)+v(i+1,j))*(u(i,j)-u(i,j+1))- &
  (v(i,j-1)+v(i+1,j-1))*(u(i,j-1)+u(i,j))- &
  gamma*abs(v(i,j-1)+v(i+1,j-1))*(u(i,j-1)-u(i,j))) & /(4.0*dely)

laplu = (u(i+1,j)-2.0*u(i,j)+u(i-1,j))/delx/delx+ &
  (u(i,j+1)-2.0*u(i,j)+u(i,j-1))/dely/dely
\end{minted}
(a) Excerpt from Fortran code for a Navier-Stoke fluid simulator,
showing highly-detailed stencil computations.
\vspace{0.5em}
\hrule
%
\begin{minted}[linenos=false,xleftmargin=1em]{fortran}
!= stencil centered(depth=1,dim=1)*reflexive(dim=2) + centered(depth=1,dim=2)*reflexive(dim=1) :: u

!= stencil forward(depth=1,dim=1) * backward(depth=1,dim=2) :: v
\end{minted}
(b) Inferred stencil specification from CamFort.
\vspace{0.5em}
\hrule
%
\begin{center}
\begin{tikzpicture}[scale=0.85]
\node at (1.25,1.22) {i, j};
\draw (1,1) rectangle (1.5,1.5);
\draw (1.5,1) rectangle (2,1.5);
\draw (1,0.5) rectangle (1.5,1);
\draw (0.5,1) rectangle (1,1.5);
\draw (1,1.5) rectangle (1.5,2);
%
\node at (4.25,1.22) {i, j};
\draw (4,1) rectangle (4.5,1.5);
\draw (4.5,1) rectangle (5,1.5);
\draw (4.5,1.5) rectangle (5,2);
\draw (4,1.5) rectangle (4.5,2);
\end{tikzpicture}
\end{center}
\vspace{-0.5em}
(c) Pictorial representation of the two stencil specifications.
The horizontal dimension is \texttt{dim=1} and the vertical is \texttt{dim=2}.
\caption{Fragment of a Navier-Stokes fluid simulator and its
  specification in our language}
\label{ref:navier-stokes-fragment}
\vspace{-0.8em}
\end{figure}

@@

In recent years, there has been an increased awareness
of the need for formal verification techniques in
science~\cite{post2005computational,oberkampf2010verification,orchard2014computational}.
Our specification language is an approach in this direction, proving
a system of lighweight specification for numerical code.
The aim is to not only provide a way of exposing errors, but also
for specifying program behaviour to future-proof the code and ease
its maintenance. In order to increase the likelihood of impact, we aim
to integrate with existing practises smoothly, hence our
specifications are integrated as comments and our tool can synthesis
specifications for the user.

Much of this paper concerns the technical aspects of the language,
its semantics, its inference, checking, and synthesis procedures.
The evaluation focuses on its application to real scientific code base
used actively in research.

@@

Our specification language is based on the observation
that most forms of array access in numerical code have
a fixed, statically-determined access pattern. For example, the
``\emph{five-point stencil}'' on a two-dimensional array reads from array
indices $(i, j)$, $(i-1, j)$, $(i+1, j)$, $(i, j-1)$, and $(i, j+1)$
for all $i, j$ within the inner boundary of the array (to avoid
out-of-bounds access at the edges). We revisit this hypothesis
in Section~\ref{sec:evaluation} where the inference of
such regular stencil patterns on a corpus of numerical programs (both
small and large).

Section~\ref{sec:syntax} outlines the syntax of specifications.
 Section~\ref{sec:eqs} provides an equational
theory for specifications via relation ($\equiv$) and a notion of
sub-specification via relation ($<:$).  Section~\ref{sec:semantics}
gives a semantic model via a set-based interpretation, for which the
equality and sub-specification relation are proven sound.

@@

\renewcommand*{\arraystretch}{0.8}
\paragraph{Target language syntax} Throughout, for the source language
syntax, $v$ ranges over program variables, $s$ over statements, and
$e$ over expressions (which may be impure).  By ``variable'', we mean
in the imperative sense (\ie{}, named binders to memory cells) instead
of mathematical variables.

@@

\begin{definition}[Checking]
Let \texttt{a} be an array variable and let
$S \in \textit{spec}$ be a specification with
the \textit{specDec} \texttt{!= stencil
  $S$ :: a} in our syntax.
For a array subscript
assignment statement $\textit{assg} = v(\overline{e_1}) = e_2$ in our source language
(with all the appropriate data flow information in place)
where $e_2$ has at most $n$-dimensional array access, 
then the assignment conforms to the specification $S$ if 
the predicate $\mathit{check}_n(S, \textit{assg}, \texttt{a})$ holds, 
defined:
\begin{align*}
& \mathit{check_n}(S, \textit{assg}, x) =
  \mathit{cons_n}(\interp{S}_n, \, \textit{modelise}(\textit{assg},
  x)) \\
& \mathit{modelise} : \textit{statement} \times \textit{variable}
  \rightarrow \textsf{Mult}((\domainVal)^n) \\ 
& \mathit{modelise} (\textit{assg}, x) =
  \begin{cases}
    \textsf{once}(\mathit{analyse}(\textit{assg})(x)) & \textit{unique}(\textit{assg},x) \\
    \textsf{mult}(\mathit{analyse}(\textit{assg})(x)) & otherwise
  \end{cases}
\end{align*}
Thus, checking proceeds by reusing the \textit{analyse} function of
inference (Section~\ref{subsec:inference}) to extract indexing
schemes for stencil computations flowing to an assignment. 
This is then augmented with multiplicity information
by $\mathit{modelise}$ where $\mathit{unique}(\textit{assg},x)$ is a
 predicate indicating whether indices to array $x$ in an
assignment $\textit{assg}$ are unique or not. 
\end{definition}

@@

\textbf{The Unified
  Model}\footnote{\url{http://www.metoffice.gov.uk/research/modelling-systems/unified-model}}
is a weather forecasting and climate modelling tool developed by the
Met Office in the United Kingdom. It is used by research organisations
and meteorological services around the world. We use the development
branch (trunk) of the model. The code based in closed source but
institutional licenses are available for research purposes. The Met Office runs a comprehensive code quality system incorporating dedicated committers (we counted 11) for particular parts of the model. We counted 120 additional contributors whose submissions are reviewed and tested before being accepted into the code base.

>>>>

\keywords{program verification, specification, stencils}

@@

In practice, most stencil computations have a regular shape that can
be described simply and abstractly with a small set of coarse-grained
spatial descriptions. We introduce a simple specification language of
such descriptions, abstracting over the fine-grained detail of stencil
access patterns. For our first Laplace example, our
inference procedure provides the following specification to line 3:
%
\begin{minted}[linenos=false]{fortran}
!=  stencil centered(depth=1, dim=1) :: a
\end{minted}
%
This describes that \texttt{a} is accessed with a
symmetrical pattern (``centered'') to a depth of one in each
direction in its first (and only) dimension.
%The second line explains
%the temporal relationship between \texttt{b} and \texttt{a}: that
%previous time step for \texttt{b} is actually provided by \texttt{a},
%and vice versa. This is explained as a mutual dependence between
%\texttt{a} and \texttt{b}.
In the Navier-Stokes example of
Figure~\ref{ref:navier-stokes-fragment}, its inferred specification is shown
in Fig.~\ref{ref:navier-stokes-fragment}(b). The
specification explains that, over the whole fragment, \texttt{u} is
read from with a centered pattern to depth of 1 in both dimensions
(this is known as the \emph{five-point stencil}) and \texttt{v}
is read from in a neighbourhood bounded by forward to depth of $1$ in
the first dimension and backward to a depth of $1$ in the second
dimension. Fig.~\ref{ref:navier-stokes-fragment}(c) illustrates
the specification pictorially.

In this paper, we make the following contributions:
%
\begin{itemize}[leftmargin=0.8em]
\item we introduce a specification language for
stencil computation that captures many common forms
of data access pattern (Section~\ref{sec:lang}), providing
 an equational theory and denotational model (Sect.~\ref{sec:semantics});

\item we detail inference, checking, and specification synthesis 
algorithms for our language (Sect.~\ref{sec:analysis}), leveraging the semantic model;

\item we evaluate our approach with an
implementation as an extension of CamFort,
an open-source program analysis tool for Fortran
programs. We use our implementation to evaluate our approach against
a corpus of scientific applications, totalling roughly one million
lines of code;

@@

\item we provide a more detailed verification case study of two
  particular stencil computation and show that we
  are able to detect the vast majority  of (synthesised) possible
  indexing errors.
\end{itemize}
%

@@

duvdy = ((v(i,j)+v(i+1,j))*(u(i,j)+u(i,j+1))+   &
  gamma*abs(v(i,j)+v(i+1,j))*(u(i,j)-u(i,j+1))- &
  (v(i,j-1)+v(i+1,j-1))*(u(i,j-1)+u(i,j))-      &
  gamma*abs(v(i,j-1)+v(i+1,j-1))*(u(i,j-1)-     & 
  u(i,j))) / (4.0*dely)

laplu = (u(i+1,j)-2.0*u(i,j)+u(i-1,j))/delx/delx+ &
  (u(i,j+1)-2.0*u(i,j)+u(i,j-1))/dely/dely
\end{minted}
(a) Excerpt of a Navier-Stokes fluid simulator
(based on~\citet{griebel1997numerical}) showing highly-detailed stencil computation.
\vspace{0.5em}
\hrule
%
\begin{minted}[linenos=false,xleftmargin=1em]{fortran}
!= stencil centered(depth=1,dim=1)*reflexive(dim=2) + centered(depth=1,dim=2)*reflexive(dim=1) :: u

!= stencil forward(depth=1,dim=1) * backward(depth=1,dim=2) :: v
\end{minted}
(b) Inferred and synthesised specification for (a) by CamFort.
\vspace{0.5em}
\hrule
%
\begin{center}
\begin{tikzpicture}[scale=0.85]
\node at (1.25,1.22) {i, j};
\draw (1,1) rectangle (1.5,1.5);
\draw (1.5,1) rectangle (2,1.5);
\draw (1,0.5) rectangle (1.5,1);
\draw (0.5,1) rectangle (1,1.5);
\draw (1,1.5) rectangle (1.5,2);
%
\node at (4.25,1.22) {i, j};
\draw (4,1) rectangle (4.5,1.5);
\draw (4.5,1) rectangle (5,1.5);
\draw (4.5,1.5) rectangle (5,2);
\draw (4,1.5) rectangle (4.5,2);
\end{tikzpicture}
\end{center}
\vspace{-0.5em}
(c) Pictorial representation of the two stencil specifications.
The horizontal dimension is \texttt{dim=1} and the vertical is \texttt{dim=2}.
\caption{Fragment of a Navier-Stokes fluid simulator and its
  specification in our language}
\label{ref:navier-stokes-fragment}
\vspace{-0.8em}
\end{figure}

@@

In recent years, there has been an increased awareness
of the need for formal verification techniques in
science~\cite{post2005computational,oberkampf2010verification,orchard2014computational}.
Our specification language is an approach in this direction, providing
a system of lighweight specification for numerical code.
The aim is to not only provide a way of exposing errors, but also
for specifying program behaviour to future-proof the code and ease
its maintenance. In order to increase the likelihood of impact, we aim
to integrate with existing practises smoothly, hence our
specifications are integrated as comments and our tool can synthesis
specifications for the user.

This paper considers the technical aspects of the language,
its semantics, inference, checking, and synthesis procedures.
We also evaluate our implementation on real scientific code 
used actively in research. Our case studies include the Unified Weather 
model used by the UK Met Office~\cite{um-website} ($\approx{}600\textit{kloc}$) and other
research-based numerical models, ranging from small to large.

@@

Our specification language is based on the observation
that most forms of array access in numerical code have
a fixed, statically-determined access pattern. For example, the
``\emph{five-point stencil}'' on a two-dimensional array reads from array
indices $(i, j)$, $(i-1, j)$, $(i+1, j)$, $(i, j-1)$, and $(i, j+1)$
for all $i, j$ within the inner boundary of the array (to avoid
out-of-bounds access at the edges). We revisit this hypothesis
in Section~\ref{sec:evaluation} with the inference of
such regular stencil patterns on our corpus of numerical programs.

Section~\ref{sec:syntax} outlines the syntax.
 Section~\ref{sec:eqs} provides an equational
theory for specifications and a notion of
sub-specification.  Section~\ref{sec:semantics}
gives a semantic model via a set-based interpretation, for which the
equality and sub-specification relation are proven sound.

@@

\renewcommand*{\arraystretch}{0.8}
\paragraph{Target language syntax} Throughout, for the source language
syntax, $e$ ranges over its expressions (which may be impure).  By ``variable'', we mean
in the imperative sense (\ie{}, named binders to memory cells) instead
of mathematical variables.

@@

\begin{definition}[Checking]
Let \texttt{a} be an array variable and let
$S \in \textit{spec}$ be a specification with
the \textit{specDec} \texttt{!= stencil
  $S$ :: a} in our syntax.
For a array subscript
assignment statement $\textit{assg} = v(\overline{e_1}) = e_2$ in our source language
(with all the appropriate data flow information in place)
where $e_2$ has at most $n$-dimensional array access, 
then the assignment conforms to the specification $S$ if 
the predicate $\mathit{check}_n(S, \textit{assg}, \texttt{a})$ holds, 
defined:
\begin{align*}
& \mathit{check_n}(S, \textit{assg}, x) =
  \mathit{cons_n}(\interp{S}_n, \, \textit{modelise}(\textit{assg},
  x)) \\
& \mathit{modelise} : \textit{statement} \times \textit{variable}
  \rightarrow \textsf{Mult}((\domainVal)^n) \\ 
& \mathit{modelise} (\textit{assg}, x) =
  \begin{cases}
    \textsf{once}(\mathit{analyse}(\textit{assg})(x)) & \textit{unique}(\textit{assg},x) \\
    \textsf{mult}(\mathit{analyse}(\textit{assg})(x)) & otherwise
  \end{cases}
\end{align*}
Thus, checking proceeds by reusing the \textit{analyse} function from 
inference (Sec.~\ref{subsec:inference}) to extract index 
schemes from source code. 
This is then augmented with multiplicity information
by $\mathit{modelise}$ where $\mathit{unique}(\textit{assg},x)$ is a
 predicate indicating whether indices to array $x$ in an
assignment $\textit{assg}$ are unique or not. 
\end{definition}

@@

\textbf{The Unified
  Model}~\cite{um-website}
is a weather forecasting and climate modelling tool developed by the
Met Office in the United Kingdom. It is used by research organisations
and meteorological services around the world. We use the development
branch (trunk) of the model. The code based in closed source but
institutional licenses are available for research purposes. The Met Office runs a comprehensive code quality system incorporating dedicated committers (we counted 11) for particular parts of the model. We counted 120 additional contributors whose submissions are reviewed and tested before being accepted into the code base.
